\texttt{nuance}: Efficient detection of planets transiting active stars

### 2024-07-19 latex/paper.tex
move figures and add arxiv submission script

% figure options
\graphicspath{{../figures}}

@@

This paper was compiled from source code available in a git-versioned repository (\href{https://github.com/lgrcia/paper-nuance}{\color{linkcolor}\faFileTextO}) and used the \textsf{snakemake}\footlink{https://snakemake.readthedocs.io} workflow management tool. Source code of the \textsf{nuance} \textsf{Python} package is also available in a git-versioned repository (\href{https://github.com/lgrcia/nuance}{\color{linkcolor}\faCode}) together with an online documentation (\href{https://nuance.readthedocs.io}{\color{linkcolor}\faBook}). Finally, we provide links to the source code used to produce each figure (inidcated by the {\color{linkcolor}\codeicon} icon).

@@

One way to quantify the detectability of a unique transit signal is to compute its signal-to-noise-ratio expressed as 
\begin{equation*}
  SNR= \frac{\Delta}{\sigma}\sqrt{n},
\end{equation*}
where $\Delta$ is the transit depth, $\sigma$ the measurements uncertainty (assuming homoscedasticity), and $n$ the number of points within transit. Although this metric is useful to assess the strength of the transit signal given a certain photometric precision, it does not account for the presence of correlated noise. However, instrumental and other astrophysical signals will necessarily affect the detectability of transits in realistic light curves \citep{Pont2006}. For this reason, the combined differential photometric precision (CDPP; \citealt{Jenkins2010}) metric was developed, and used in the context of the Kepler mission to assess the level of correlated noise in light curves, affecting the detectability of transits with a given duration. 
% Ultimately, this metric was essential to estimate the yield of the survey, and helped to achieve one of the primary goal of the mission: to measure the occurrence rate of Earth-size planets orbiting in the habitable zones of sun-like stars \citep{}.
The CDPP is computed by decomposing the data in the time-frequency domain using wavelets, and measures the significance of a distorted transit signal in the whitened data. For a given transit duration, the CDPP is then a measure of the noise remaining after filtering the light curve in each frequency band, taking into account the presence of nonstationary correlated noise of a given timescale (see \citealt{Jenkins2010} for more details). Hence, we can estimate the significance of the transit signal accounting for the presence of correlated noise as
\begin{equation}\label{eq:snr}
  SNR= \frac{\hat{\Delta}}{CDPP_D},
\end{equation}
where $CDPP_D$ is the CDPP computed for a given transit duration $D$ and $\hat{\Delta}$ is the transit depth after detrending. For simplicity, the CDPP is computed using the method from \cite{Gilliland2011} implemented in the \textsf{lightkurve} Python package \citep{Lightkurve} and $\hat{\Delta}$ is the depth obtained by solving \autoref{eq:linear_model} with the last column of $\bm{X}$ containing a normalized transit model.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.6\linewidth]{../workflows/plot_issues/figures/issue1.pdf}
        \caption{Illustration of the effect of correlated noise on a single transit SNR. An hour-long transit of depth 0.5\% is generated on top of white noise (standard deviation of 0.15\%) as part of a 24-hours observation with an exposure time of 1 minute (top). Then, in the bottom plot, correlated noise is added, generated using a GP with an SHO kernel of hyperparameters $\omega=60$, $Q=0.5$ and $\sigma=\Delta/4$ . The SNR on the right of each light curve is computed using \autoref{eq:snr}. Models used to simulate these data are provided in \autoref{light_curves_simulations}. \codelink{workflows/plot_issues/scripts/plot_issue_1.py}}
        \label{fig:issue1}
    \end{centering}
\end{figure}
\autoref{fig:issue1} shows the SNR from \autoref{eq:snr} computed for a unique transit observed in the absence (gray) and presence of correlated noise (red). As illustrated in this figure, the presence of correlated noise strongly decreases the transit signal SNR, which would ultimately limit its detectability.

@@

        \includegraphics[width=\linewidth]{../workflows/plot_issues/figures/issue2.pdf}

@@

        \includegraphics[width=\linewidth]{../workflows/cleaning_snr/figures/result.pdf}

@@

The goal of the linear search is to compute the likelihood $p(\bm{f} \vert T_0 , D, \Delta)$ for a grid of epochs $\set{T_i}_i$ and durations $\set{D_j}_j$. For each pair $(T_i, D_j)$ the transit depth $\Delta_{i, j}$ is linearly solved, which leads to the set of maximum likelihoods
\begin{equation*}
    \set{\ln\mathcal{L}_{i,j}}_{i, j} = \set{\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i, j})}_{i, j}.
\end{equation*}
An example of such a grid of likelihoods is shown in \autoref{fig:linear_search}. To prepare for the next step, uncertainties on the depths $\sigma_{i,j}$ are also computed using \autoref{eq:ls} and stored.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/principle/figures/principle_linear_search.pdf}
        \caption{Principle and output of the linear search. The simulated dataset (top) corresponds to the one shown and described in \autoref{fig:app_principle_dataset}. First, a set of durations and depths $\set{T_i, D_j}_{i,j}$ is generated. For each pair of indices $(i,j)$, the likelihood $\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})$ is computed using the parameters from \autoref{eq:ls} and the expression of \autoref{eq:linear_search_ll}. This process yields the grid of log-likelihoods $\ln\mathcal{L}$ (bottom plot), as well as the $\set{\Delta_{i,j}, \sigma_{i,j}}_{i, j}$ transit depths and errors inferred linearly using \autoref{eq:ls}. \codelink{workflows/principle/principle.ipynb}}
        \label{fig:linear_search}
    \end{centering}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{../workflows/principle/figures/principle_periodic_1.pdf}\\
    \caption{Periodic search likelihood $\mathcal{P}(P)$ computed for different trial periods $P$. Notice how the maximum value of $\mathcal{P}$ for the alias period $P=0.35$ day (left plot) is lower than for $P=2\times 0.35=0.7$ day, a result of combining the log-likelihoods using \autoref{eq:result} instead of \autoref{eq:attempt}, in favor of individual transits matching a common depth $\Delta$. \codelink{workflows/principle/principle.ipynb}}
    \label{fig:periodic_search}
\end{figure}

@@

Using \autoref{eq:result}, we can now compute $\ln\mathcal{P}$ for a range of periods and phases, and build a transit search periodogram using \autoref{eq:periodogram}. This has two disadvantages: First, each likelihood $p(\bm{f} \vert T_0, D, \Delta)$ estimated during the linear search is computed using N measurements. Hence, combining transits in the periodic search, through $\Delta_k$, $\sigma_k$ and the product of $K$ likelihoods $\set{\mathcal{L}_k}_k$ (see \autoref{eq:result}), artificially leads to a likelihood involving up to $N\times K$ measurements. For this reason, one has to normalize each likelihood $\ln\mathcal{P}$, by keeping track of the number of points used to compute each of them, which differs from one phase to another. Second, the maximum value of the likelihood $\ln\mathcal{P}$ is relative to a given dataset, so that a more intuitive and absolute metric must be used to relate to the transit signal detection (such as the Signal Detection Efficiency in \citealt{bls}). This motivates a final step to produce an interpretable transit search periodogram $\mathcal{Q}$. 
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/principle/figures/principle_Q.pdf}
        \caption{For each period $P$, the joint likelihood $\mathcal{P}(P)$ is computed using \autoref{eq:result}, and the value of the maximum likelihood transit SNR retained as $\mathcal{Q}(P)$. \codelink{workflows/principle/principle.ipynb}}
        \label{fig:periodogram}
    \end{centering}
\end{figure}
For any period P, instead of taking $ \mathcal{Q}(P)$ as the maximum value of $\ln\mathcal{P}$, we compute the maximum likelihood parameters
\begin{equation}\label{eq:phi0}
    (\phi_0 ,D) = \argmax_{\phi_i, D_j} \set{\ln p(\bm{f} \vert P, \phi_i, D_j)}_{i, j}
\end{equation}
and define $\mathcal{Q}(P)$ as being equal to the SNR of the transit of period $P$, epoch $T_0 = \phi_0 \times P$, duration $D$ and depth $\Delta$, i.e.
\begin{equation*}
    \mathcal{Q}(P) = \frac{\Delta}{\sigma},
\end{equation*}
where $\Delta$ and $\sigma$ are obtained using \autoref{eq:result} with the last column of $X$ containing a periodic transit signal of period $P$, epoch $T_0$, duration $D$ and depth $1$. This process and the resulting periodogram $\mathcal{Q}$ are shown in \autoref{fig:periodogram}.\\\\
Hence, periodic transit of period $P$ with the maximum SNR, i.e. maximizing $\mathcal{Q}$, is adopted as the best candidate, basing the confidence in this signal through its SNR. The parameters of this transit are the period $P$, epoch $T_0 = \phi_0 P$, duration $D$ (\autoref{eq:phi0}), and depth $\Delta$ with error $\sigma$ (given by \autoref{eq:result}).\\\\
\subsection{An open-source python package}\label{package}
The methods presented in this paper are made available through the \nuancecode{} open-source Python package hosted at \href{https://github.com/lgrcia/nuance}{https://github.com/lgrcia/nuance}, released on the Python Package Index\footlink{https://pypi.org/project/nuance/} and with documentation and tutorials hosted at \href{https://nuance.readthedocs.io}{https://nuance.readthedocs.io}. All following mentions of \nuancecode{} refer to its \review{version 0.6.0 \citep{zenodo}.}{} \\\\
To instantiate a search, a user can start by creating a \textsf{Nuance} object with
\begin{lstlisting}[language=Python]
from nuance import Nuance

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/control_test_bls/figures/control_test.pdf}
        \caption{Binned statistics of the injection-recovery of 4000 transit signals in a flat light curve with only white noise using \textsf{BLS} and \nuancecode{}. The color scale indicates the recovery of transits in the corresponding (period, depth) parameter space, white for a full recovery and black for no detection. \codelink{workflows/control_test_bls/scripts/plot.py}}
        \label{fig:control}
    \end{centering}
\end{figure}

@@

        \includegraphics[width=\linewidth]{../workflows/synthetic-injection-recovery/figures/synthetic_ir.pdf}

@@

        \includegraphics[width=\linewidth]{../workflows/tess_injection_recovery/figures/tpfp_bars.pdf}

@@

        \includegraphics[width=\linewidth]{../workflows/tess_injection_recovery/figures/tpfp.pdf}

@@

        \includegraphics[width=\linewidth]{../workflows/tess_injection_recovery/figures/tpfp_ims.pdf}

@@

        \includegraphics[width=0.6\linewidth]{../workflows/tess_injection_recovery/figures/tpfp_radius.pdf}

@@

        \includegraphics[width=\linewidth]{../workflows/comparison_toi/figures/TOI 540/4.pdf}

@@

        \includegraphics[width=\linewidth]{../workflows/comparison_toi/figures/TOI 540/periodograms.pdf} 

@@

\noindent In \autoref{fig:blsvsnuance}, the processing time of \nuancecode{} linear and periodic search are recorded against the number of points in a simulated light curve, assuming a simple un-optimized GP with a squared exponential kernel. These are compared to the processing time of \texttt{bi-weight+BLS} (see \autoref{simu}) separated into the bi-weight filtering step and the BLS search.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/benchmark/figures/nuance_vs_bls.pdf}
        \caption{Study of \textsf{nuance} (black line) processing time against \texttt{bi-weight+BLS} (see \autoref{simu}, blue line). The gray curve shows the performance of \textsf{nuance} linear search when applied to chunks of 10 000 points continuous observations, instead of considering these observations all together. This study is performed on a single CPU core of an Apple M2 Max chip. While not being shown, we verify that both the BLS algorithm and \nuancemethod{} processing times scale linearly with the number of trial transit durations and orbital periods. \codelink{workflows/benchmark/scripts/plot.py}}
        \label{fig:blsvsnuance}
    \end{centering}
\end{figure}

>>>>

% figure options
\graphicspath{{./figures}}

@@

This paper was compiled from source code available in a git-versioned repository (\href{https://github.com/lgrcia/paper-nuance}{\color{linkcolor}\faFileTextO}) and used the \textsf{snakemake}\footlink{https://snakemake.readthedocs.io} workflow management tool. Source code of the \textsf{nuance} \textsf{Python} package is also available in a git-versioned repository (\href{https://github.com/lgrcia/nuance}{\color{linkcolor}\faCode}) together with an online documentation (\href{https://nuance.readthedocs.io}{\color{linkcolor}\faBook}). Finally, we provide links to the source codes used to produce each figure (indicated by {\color{linkcolor}\codeicon}).

@@

One way to quantify the detectability of a unique transit signal is to compute its signal-to-noise-ratio expressed as 
\begin{equation*}
  SNR= \frac{\Delta}{\sigma}\sqrt{n},
\end{equation*}
where $\Delta$ is the transit depth, $\sigma$ the measurements uncertainty (assuming homoscedasticity), and $n$ the number of points within transit. Although this metric is useful to assess the strength of the transit signal given a certain photometric precision, it does not account for the presence of correlated noise. However, instrumental and other astrophysical signals will necessarily affect the detectability of transits in realistic light curves \citep{Pont2006}. For this reason, the combined differential photometric precision (CDPP; \citealt{Jenkins2010}) metric was developed, and used in the context of the Kepler mission to assess the level of correlated noise in light curves, affecting the detectability of transits with a given duration. 
% Ultimately, this metric was essential to estimate the yield of the survey, and helped to achieve one of the primary goal of the mission: to measure the occurrence rate of Earth-size planets orbiting in the habitable zones of sun-like stars \citep{}.
The CDPP is computed by decomposing the data in the time-frequency domain using wavelets, and measures the significance of a distorted transit signal in the whitened data. For a given transit duration, the CDPP is then a measure of the noise remaining after filtering the light curve in each frequency band, taking into account the presence of nonstationary correlated noise of a given timescale (see \citealt{Jenkins2010} for more details). Hence, we can estimate the significance of the transit signal accounting for the presence of correlated noise as
\begin{equation}\label{eq:snr}
  SNR= \frac{\hat{\Delta}}{CDPP_D},
\end{equation}
where $CDPP_D$ is the CDPP computed for a given transit duration $D$ and $\hat{\Delta}$ is the transit depth after detrending. For simplicity, the CDPP is computed using the method from \cite{Gilliland2011} implemented in the \textsf{lightkurve} Python package \citep{Lightkurve} and $\hat{\Delta}$ is the depth obtained by solving \autoref{eq:linear_model} with the last column of $\bm{X}$ containing a normalized transit model.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.6\linewidth]{./issue1.pdf}
        \caption{Illustration of the effect of correlated noise on a single transit SNR. An hour-long transit of depth 0.5\% is generated on top of white noise (standard deviation of 0.15\%) as part of a 24-hours observation with an exposure time of 1 minute (top). Then, in the bottom plot, correlated noise is added, generated using a GP with an SHO kernel of hyperparameters $\omega=60$, $Q=0.5$ and $\sigma=\Delta/4$ . The SNR on the right of each light curve is computed using \autoref{eq:snr}. Models used to simulate these data are provided in \autoref{light_curves_simulations}. \codelink{workflows/plot_issues/scripts/plot_issue_1.py}}
        \label{fig:issue1}
    \end{centering}
\end{figure}
\autoref{fig:issue1} shows the SNR from \autoref{eq:snr} computed for a unique transit observed in the absence (gray) and presence of correlated noise (red). As illustrated in this figure, the presence of correlated noise strongly decreases the transit signal SNR, which would ultimately limit its detectability.

@@

        \includegraphics[width=\linewidth]{./issue2.pdf}

@@

        \includegraphics[width=\linewidth]{./result.pdf}

@@

The goal of the linear search is to compute the likelihood $p(\bm{f} \vert T_0 , D, \Delta)$ for a grid of epochs $\set{T_i}_i$ and durations $\set{D_j}_j$. For each pair $(T_i, D_j)$ the transit depth $\Delta_{i, j}$ is linearly solved, which leads to the set of maximum likelihoods
\begin{equation*}
    \set{\ln\mathcal{L}_{i,j}}_{i, j} = \set{\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i, j})}_{i, j}.
\end{equation*}
An example of such a grid of likelihoods is shown in \autoref{fig:linear_search}. To prepare for the next step, uncertainties on the depths $\sigma_{i,j}$ are also computed using \autoref{eq:ls} and stored.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{./principle_linear_search.pdf}
        \caption{Principle and output of the linear search. The simulated dataset (top) corresponds to the one shown and described in \autoref{fig:app_principle_dataset}. First, a set of durations and depths $\set{T_i, D_j}_{i,j}$ is generated. For each pair of indices $(i,j)$, the likelihood $\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})$ is computed using the parameters from \autoref{eq:ls} and the expression of \autoref{eq:linear_search_ll}. This process yields the grid of log-likelihoods $\ln\mathcal{L}$ (bottom plot), as well as the $\set{\Delta_{i,j}, \sigma_{i,j}}_{i, j}$ transit depths and errors inferred linearly using \autoref{eq:ls}. \codelink{workflows/principle/principle.ipynb}}
        \label{fig:linear_search}
    \end{centering}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{./principle_periodic_1.pdf}\\
    \caption{Periodic search likelihood $\mathcal{P}(P)$ computed for different trial periods $P$. Notice how the maximum value of $\mathcal{P}$ for the alias period $P=0.35$ day (left plot) is lower than for $P=2\times 0.35=0.7$ day, a result of combining the log-likelihoods using \autoref{eq:result} instead of \autoref{eq:attempt}, in favor of individual transits matching a common depth $\Delta$. \codelink{workflows/principle/principle.ipynb}}
    \label{fig:periodic_search}
\end{figure}

@@

Using \autoref{eq:result}, we can now compute $\ln\mathcal{P}$ for a range of periods and phases, and build a transit search periodogram using \autoref{eq:periodogram}. This has two disadvantages: First, each likelihood $p(\bm{f} \vert T_0, D, \Delta)$ estimated during the linear search is computed using N measurements. Hence, combining transits in the periodic search, through $\Delta_k$, $\sigma_k$ and the product of $K$ likelihoods $\set{\mathcal{L}_k}_k$ (see \autoref{eq:result}), artificially leads to a likelihood involving up to $N\times K$ measurements. For this reason, one has to normalize each likelihood $\ln\mathcal{P}$, by keeping track of the number of points used to compute each of them, which differs from one phase to another. Second, the maximum value of the likelihood $\ln\mathcal{P}$ is relative to a given dataset, so that a more intuitive and absolute metric must be used to relate to the transit signal detection (such as the Signal Detection Efficiency in \citealt{bls}). This motivates a final step to produce an interpretable transit search periodogram $\mathcal{Q}$. 
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{./principle_Q.pdf}
        \caption{For each period $P$, the joint likelihood $\mathcal{P}(P)$ is computed using \autoref{eq:result}, and the value of the maximum likelihood transit SNR retained as $\mathcal{Q}(P)$. \codelink{workflows/principle/principle.ipynb}}
        \label{fig:periodogram}
    \end{centering}
\end{figure}
For any period P, instead of taking $ \mathcal{Q}(P)$ as the maximum value of $\ln\mathcal{P}$, we compute the maximum likelihood parameters
\begin{equation}\label{eq:phi0}
    (\phi_0 ,D) = \argmax_{\phi_i, D_j} \set{\ln p(\bm{f} \vert P, \phi_i, D_j)}_{i, j}
\end{equation}
and define $\mathcal{Q}(P)$ as being equal to the SNR of the transit of period $P$, epoch $T_0 = \phi_0 \times P$, duration $D$ and depth $\Delta$, i.e.
\begin{equation*}
    \mathcal{Q}(P) = \frac{\Delta}{\sigma},
\end{equation*}
where $\Delta$ and $\sigma$ are obtained using \autoref{eq:result} with the last column of $X$ containing a periodic transit signal of period $P$, epoch $T_0$, duration $D$ and depth $1$. This process and the resulting periodogram $\mathcal{Q}$ are shown in \autoref{fig:periodogram}.\\\\
Hence, periodic transit of period $P$ with the maximum SNR, i.e. maximizing $\mathcal{Q}$, is adopted as the best candidate, basing the confidence in this signal through its SNR. The parameters of this transit are the period $P$, epoch $T_0 = \phi_0 P$, duration $D$ (\autoref{eq:phi0}), and depth $\Delta$ with error $\sigma$ (given by \autoref{eq:result}).\\\\
\subsection{An open-source python package}\label{package}
The methods presented in this paper are made available through the \nuancecode{} open-source Python package hosted at \href{https://github.com/lgrcia/nuance}{https://github.com/lgrcia/nuance}, released on the Python Package Index\footlink{https://pypi.org/project/nuance/} and with documentation and tutorials hosted at \href{https://nuance.readthedocs.io}{https://nuance.readthedocs.io}. All following mentions of \nuancecode{} refer to its \review{version 0.6.0 \citep{zenodo}.}{} \\\\
To instantiate a search, a user can start by creating a \textsf{Nuance} object with
\begin{lstlisting}[language=Python]
from nuance import Nuance

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{./control_test.pdf}
        \caption{Binned statistics of the injection-recovery of 4000 transit signals in a flat light curve with only white noise using \textsf{BLS} and \nuancecode{}. The color scale indicates the recovery of transits in the corresponding (period, depth) parameter space, white for a full recovery and black for no detection. \codelink{workflows/control_test_bls/scripts/plot.py}}
        \label{fig:control}
    \end{centering}
\end{figure}

@@

        \includegraphics[width=\linewidth]{./synthetic_ir.pdf}

@@

        \includegraphics[width=\linewidth]{./tpfp_bars.pdf}

@@

        \includegraphics[width=\linewidth]{./tpfp.pdf}

@@

        \includegraphics[width=\linewidth]{./tpfp_ims.pdf}

@@

        \includegraphics[width=0.6\linewidth]{./tpfp_radius.pdf}

@@

        \includegraphics[width=\linewidth]{./TOI_540/4.pdf}

@@

        \includegraphics[width=\linewidth]{./TOI_540/periodograms.pdf} 

@@

\noindent In \autoref{fig:blsvsnuance}, the processing time of \nuancecode{} linear and periodic search are recorded against the number of points in a simulated light curve, assuming a simple un-optimized GP with a squared exponential kernel. These are compared to the processing time of \texttt{bi-weight+BLS} (see \autoref{simu}) separated into the bi-weight filtering step and the BLS search.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{./nuance_vs_bls.pdf}
        \caption{Study of \textsf{nuance} (black line) processing time against \texttt{bi-weight+BLS} (see \autoref{simu}, blue line). The gray curve shows the performance of \textsf{nuance} linear search when applied to chunks of 10 000 points continuous observations, instead of considering these observations all together. This study is performed on a single CPU core of an Apple M2 Max chip. While not being shown, we verify that both the BLS algorithm and \nuancemethod{} processing times scale linearly with the number of trial transit durations and orbital periods. \codelink{workflows/benchmark/scripts/plot.py}}
        \label{fig:blsvsnuance}
    \end{centering}
\end{figure}
